Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data

### 2014-04-09 image_analysis_paper/pnas_draft.tex
Added more edits

\title{Temporal ordering and registration of imaging data in cross--sectional studies of development}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal \cite{...} and cross-sectional \cite{...}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory; the latter must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered")  and ordered in time.
%
We present a general computational framework that greatly accelerates -and can combine- both of these tasks.

@@

%
To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of regulatory protein expression 
%%%
%%% check whether "regulatory protein" is the correct term
%%%
pattern formation in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now rotationally registered and temporally ordered. For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and an azimuthally nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

@@

The task of image registration has been widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that, in most of these applications, registration first requires defining a set of {\em features} or landmark points, and then locating these features within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then compare images by comparing these features.
%
Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}; these methods are also based on first extracting features from the images.
%
However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}; there each image is a picture of (essentially) the {\em same} molecule, and variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics (which underpins the development of vector diffusion map algorithms).

@@

%%%   you should sample well enough in time
%%% your dmap coordinate should have a nice slope with time
%% you should have variation in angle ---and we know we are random in the rotations
%%
%% and now the green is an issue of noise.....
%% the green "sees" angle but not time,
%% the red sees angle and time, but angle badly
%% the ocmbination sees both well enough 
%%
%%
%%   we need to say things about "other methods" -- Mallat, AmitNow, Procrustes...
%% and if we change the artificial stuff, but even if not
%% we need to say something about flips.
%%
%%   Y is afraid of the compsci literature being "fairly represented" 
%%
%%

@@

Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ pertains to the diffusion kernel for the pair of data points $x_i, x_j$ and is defined in \eqref{eq:dmaps_A}, and $H_{ij} \in \mathbb{R}^{d \times d}$ pertains to the pairwise alignment between the data points and is defined in \eqref{eq:H_to_R}.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment; that is, the minimum obtained in \eqref{eq:opt_pairwise} among all possible distances between all possible shifts of the two data points.
%
In the language of symmetry groups, this distance is a metric between the orbits of the data points induced by the symmetry group.

@@

Registering images with respect to rotations {\em and translations} requires some additional care.
%
%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
%
For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$.
%
However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
%
Instead, we (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices, by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
%
Rotation of the image corresponds to rotation around one principal axis in three dimensional, and translation of the image corresponds (approximately) to rotations around the other two principal axes. %(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
%
%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

To register the images, we first compute the translations and rotations to optimally align each pair of data points.
%
Practically, the images are discretized at pixels (rather than continuous functions on the unit sphere).
%
We therefore discretize the optimization problem, and for each image pair $I_i$ and $I_j$, we solve
\begin{equation}\label{eq:opt_pairwise}
(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
\scriptsize \begin{matrix}
\theta \in [0, 2\pi) \\
dx, dy \in [-\Delta, \Delta]\\
%dy \in [-\Delta, \Delta]
\end{matrix}
} \|g(I_j, \theta, dx, dy) - I_i \|^2.
\end{equation}
where $\Delta$ is the maximum number of pixels by which we translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
%
The norm we use, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
%
The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
%
The missing pixels in the rotated image due to corner effects are taken to be black.
%
The translations are implemented by shifting the pixels with periodic boundary conditions.
%
In \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).
%
We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to estimate the solution.
%
We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
and 11 possible translations in both the $x$ and $y$ directions. %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
We then check all possible combinations for the best rotation and translations that align $I_j$ to $I_i$.
%
This direct enumeration approach is not prohibitive for this step.
%
%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
%
%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
%
%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.

%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
%
%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
%
We then convert the two-dimensional translation and rotation to a (continuous) three-dimensional rotation.
%
We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
%
This portion of the sphere is small enough so that curvature effects are not seen in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

From the rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding translations and rotations of the images.
%
We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
%
From a rotation matrix $R$, we can compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) &
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) &
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta)  &
\end{aligned}
\end{equation}
%
We then invert \eqref{eq:angle_relations} to compute the corresponding translation and rotation.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.

>>>>

\title{Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory; the latter must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered") to factor out the relevant symmetries (translations and rotations),  and ordered in time.
%
We present a general computational framework that greatly accelerates -and can combine- both of these tasks.

@@

%
To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of regulatory protein expression
%%%
%%% check whether "regulatory protein" is the correct term
%%%
pattern formation in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now rotationally registered and temporally ordered. For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}
%
The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and an azimuthally nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

@@

The task of image registration has been widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
The main point to note here is that many registration algorithms require the locations of landmarks or features within each image.
%
For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then compare images by comparing these features.
%
Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}; these methods are also based on first extracting features from the images.
%
However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
%
Template methods are also common for registration, where one optimally aligns each image to a predefined template function.
%
However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
%
Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}; there each image is a picture of (essentially) the {\em same} molecule, and variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
%
In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics (which underpins the development of vector diffusion map algorithms).
%
Other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
%
These synchronization methods are formulated for general symmetry groups.
%
Depending on the specific data, one might consider other symmetries, such as reflections and scaling/dilation.
%
For synchronization, our symmetry group must be such that...

Temporal ordering can also be accomplished without first registering the images, by constructing {\em features} which are invariant to the underlying symmetry group. 
%
In such methods, one computes functions of the images which are invariant under translations and rotations. 
%
These functions are then used as the data in the diffusion maps calculation; since the functions do not change under translations and rotations, the resulting diffusion maps embedding is invariant to symmetries.
%
Such functions include the scattering transform \cite{mallat2012group} and the bispectrum \cite{zhao2014rotationally}.
%
These have the disadvantage that a we do not obtain a registered set of images, but only a temporal ordering of the images.
%
However, such methods can often be faster for large data sets, since sets of features are computed per image, whereas synchronization requires pairwise computations.

@@

%%%   you should sample well enough in time
%%% your dmap coordinate should have a nice slope with time
%% you should have variation in angle ---and we know we are random in the rotations
%%
%% and now the green is an issue of noise.....
%% the green "sees" angle but not time,
%% the red sees angle and time, but angle badly
%% the ocmbination sees both well enough
%%
%%
%%   we need to say things about "other methods" -- Mallat, AmitNow, Procrustes...
%% and if we change the artificial stuff, but even if not
%% we need to say something about flips.
%% we need to say symmetry early at the beginning -- important
%%  the expression "factoring out symmetry", the reference to Ben....  something about symmetry groups 
%%    the smoothness when one wants to make movies.
%%  
%%
%%   Y is afraid of the compsci literature being "fairly represented"
%%
%%

@@

Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ pertains to the diffusion kernel for the pair of data points $x_i, x_j$ and is defined in \eqref{eq:dmaps_A}, and $H_{ij} \in \mathbb{R}^{d \times d}$ pertains to the pairwise alignment between the data points and is defined in \eqref{eq:H_to_R}.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment; that is, the minimum obtained in \eqref{eq:opt_pairwise} among all possible distances between all possible shifts of the two data points.
%
In the language of symmetry groups, this distance is a metric between the orbits of the data points induced by the symmetry group.

@@

Registering images with respect to rotations {\em and translations} requires some additional care.
%
%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
%
%We first compute the translations and rotations to optimally align each pair of data points.
%
Practically, we have planar square images discretized as pixels (rather than continuous functions on the plane).
%
We therefore solve discrete pairwise optimization problems: for each image pair $I_i$ and $I_j$ we find
\begin{equation}\label{eq:opt_pairwise}
(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
\scriptsize \begin{matrix}
\theta \in [0, 2\pi) \\
dx, dy \in [-\Delta, \Delta]\\
%dy \in [-\Delta, \Delta]
\end{matrix}
} \|g(I_j, \theta, dx, dy) - I_i \|^2.
\end{equation}
where $g(I_j, \theta, dx, dy)$ is a function that results from first rotating the image $I_j$ around the center of the square by $\theta$ degrees, then translating the resulting
image by $dx$ pixel widths in the $x$ direction, and finally translating by $dy$ pixel widths in the $y$ direction.
%
The norm we use, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
%
The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
%
This also explains the $\Delta$ limits in the translation: 
%
in \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image); it does not make sense to shift the images so much that
the embryo would ``move out" of the original square.

%
Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
%
The missing pixels in the rotated image (the ``corner effects" we mentioned above) are taken to be black (zero intensity).
%
The translations are implemented by shifting the pixels (for computational speed we used periodic boundary conditions).
%
We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
%
Our ``quantum translation" is chosen to be $1/10^{th}$ of the domain in each of the $x$ and $y$ directions, while
our ``quantum rotation" is chosen to be $1/20^{th}$ of $2 \pi$. 
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
We then check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive.
%
%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
%
%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
%
%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.

%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
%
%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
%

The requirements for the synchronization step (e.g. for \eqref{eq:H_low_rank} to be satisfied), include the condition that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$ \cite{singer2013spectral}.
%
However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
%
Instead, we {\em approximately} represent rotations and translations in two-dimensions using {\em three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
%
Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes. %(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
%
%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

We thus convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations.
%
We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
%
This portion of the sphere is small enough so that curvature effects are not prominent in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps, to compute the estimates of the $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding (approximately) optimal translations and rotations of the images.
%
We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
%
From a rotation matrix $R$, we can compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) &
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) &
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta)  &
\end{aligned}
\end{equation}
%
Inverting \eqref{eq:angle_relations} yields the corresponding approximate translations and rotation;
optimal translations to the nearest integer to remove the need for interpolation when translating the images.
